### 2019-09-22 research/paper AGM/paper_body.tex
Add FG paper format

The field of affective computing studies the development of systems that can recognize, interpret, process, and simulate human affects. A subfield of affective computing that we are interested in is speech emotion recognition (SER). SER is the problem of recognizing which emotions are present in speech and SER has been approached in two ways:
\begin{itemize}
	\item Feature engineering
	\item Machine learning (ML)
\end{itemize}

@@

Feature engineering is a method of manually extracting desired parameters from the data for use in  recognition. For audio, examples of parameters include pitch, energy, and Mel-Frequency Cepstral Coefficients (MFCC).\cite{Rybka2013} Nassif et al. \cite{Nassif2019} found that most researchers used the MFCC parameter for deep learning models but also recommended other parameters such as Linear Predictive Coding.

Machine learning is a method of creating an artificial neural network that automatically extracts its own parameters for recognition. Examples include support vector machines (SVM), convolutional neural networks (CNN), and recurrent neural networks (RNN). Both approaches are not exclusive and a hybrid approach of feature engineering and machine learning results in better classification accuracy.\cite{Nassif2019}

@@

Current state-of-the-art SER has achieved an accuracy of 52.14\% for the case of speaker-independent, single-labeled emotions on the IEMOCAP database.\cite{Zhao2019} This result was achieved using a 2D CNN long short-term memory (LSTM) model where the audio waveform was converted into a log-Mel spectrogram and then fed into the model.

@@

The IEMOCAP \cite{busso_2008} and CREMA-D \cite{cao_2014} databases both include multiple labels for each speech sample but the data is discarded by considering the emotion with the majority of votes. We argue that discarding the non-majority votes results in a less realistic model of emotion classification due to not matching human performance and due to the loss of information by discarding votes.

@@

We approach the problem of multi-label emotion classification by using a 2D CNN to classify speech samples into multiple emotions. We collected four databases: two with multi-label and single-label samples, and two with only single-label samples. We then combined the four databases into an a single database by processing all of the speech samples into log-Mel spectrograms. These log-Mel spectrograms were then fed into an eight-layer neural network consisting of four convolutional layers and four dense layers.

This paper is outlined as follows:
\begin{itemize}
	\item Section II details the methodology we use such as the how we combined the four databases and describes the architecture of the neural network.
	\item Section III describes and discusses the results from testing the neural network.
	\item Section IV concludes this paper and provides suggestions for future directions.
\end{itemize}

@@

We approach the problem of SER by preprocessing samples from four databases and then feeding those samples into a CNN. We preprocess the speech samples into log-Mel spectrogram to help the neural network extract features relevant to emotion recognition. This choice is based on previous work such as \cite{Engel2019} and \cite{Badshah2017}. The choice of using a CNN is also based on previous work where Balakrishnan et al. \cite{Balakrishnan2017} found that CNN-based models have superior performance compared to RNNs. Another justification for using CNNs is due treating the log-Mel spectrograms as images which CNNs have been shown to perform well on.\cite{Krizhevsky2012}

A high level overview of the data flow is shown in figure \ref{highLevelDataFlowDiagram} where samples flow through four stages of processing. The stages are detailed further into the methodology. Both the preprocessing steps and the CNN were implemented in Python using the Librosa \cite{McFee2015} and Keras \cite{Chollet2015} library respectively.

\subsection{Preprocessing}

@@

To determine if a ML model can perform multi-label speech emotion recognition, we first need to obtain speech samples with their labeled emotion. We considered eight databases and chose four based on accessibility and the number of overlapping emotions. Table \ref{dbEmoTable} compares the set of emotions of each database which makes it easier to identify overlapping emotions. We chose four databases because it becomes more difficult to maintain consistency due to database variability. Databases can vary in
\begin{itemize}
	\item The set and number of labeled emotions
	\item The number of labels per sample
	\item The audio quality such as sampling rate and noise
	\item The spoken language
\end{itemize}
Given this variability, we chose the following four databases
\begin{enumerate}
	\item IEMOCAP \cite{busso_2008}
	\item TESS \cite{dupuis_2011}
	\item RAVDESS \cite{livingstone_2018}
	\item CREMA-D \cite{cao_2014}
\end{enumerate}

@@

To control for the number of labels per sample, we mixed the multi-labeled data from the IEMOCAP and CREMA-D databases with the single-labeled data from the TESS and RAVDESS databases. We chose to mix of single- and multi- labeled data to increase the number of samples that the model can learn from and because the single-labels can be considered as special cases of multi-labeled data. However, we removed samples that were labeled with four or five emotions because we consider samples with that many emotions to be ambiguous and because outliers hinder a neural networks ability to learn. Outliers can cause large gradient updates which prevents a model from converging.

@@

Each sample from a database would flow through the same preprocessing steps to maintain consistency across all samples. The detailed steps are described below.
\begin{enumerate}
	\item A sample starts as a raw waveform in the form of time series points specifying the amplitude at a certain point in time.
	\item The sample is then padded or cropped to the desired length of 4.5s. Shorter samples were zero-padded on the right tail. Longer samples were cropped and the extra information discarded.
	\item If the sample came from a database that we considered noisy, then a noise reduction filter was applied to the sample. We consider the IEMOCAP and CREMA-D databases to be noisy.
	\item The sample is then converted into a log-Mel spectrogram using the short-time Fourier transform (STFT) and Mel scale equation. The phase information was discard as it does not seem to hold relevant information. \cite{Kozakowski2017}
	\item The final step is to normalize the spectrograms to have values between negative one and one. This was done by using a min-max scaling function.
\end{enumerate}

The discrete-time STFT is defined in (\ref{stft}) where $x[n]$ is the discrete signal and $w[n]$ is the window function.

@@

The Mel scale is defined in (\ref{melScale}) where $f$ is the frequency in hertz and $m$ is mels.\cite{Oshaughnessy1990}

@@

For the STFT, we used a window size of 3072 with a 75\% overlap. This choice is based on the work of Zhao et al. \cite{Zhao2019} where they also use 75\% overlap but with a smaller window size (2048) and they achieved excellent results. For the Mel spectrogram, we set the minimum frequency to 20 Hz and maximum frequency to 12 kHz with 200 Mel bins. The frequency range was chosen after experimenting with various ranges and selecting the range that resulted in visually clean spectrograms. The number of Mel bins was also chosen after experimentation as too few bins resulted in poor temporal resolution while too many bins resulted in poor frequency resolution. This is due to the Gabor limit which describes the trade-off between temporal and frequency resolution.

@@

\begin{figure*}
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{neural_network_architecture.png}
	\caption{Architecture for the speech emotion recognition network including the input layer and eight processing layers.}
	\label{neuralNetworkArchitecture}
\end{figure*}

After all of the databases were processed this way, the final combined database was fed into a neural network for training, validation, and testing. We constructed an eight layer neural network consisting of four convolutional layers and three dense layers. The network architecture is shown in figure \ref{neuralNetworkArchitecture}. The training process is described below.
\begin{itemize}
	\item We shuffled the combined database to make each input batch more uniform thus mitigating large gradient updates.
	\item We split the combined database into 80\% training, 10\% validation, and 10\% testing.
	\item We applied dropout to the dense layers and batch normalization to the convolutional layers to deal with overfitting. \cite{Srivastava2014}, \cite{Ioffe2015}
	\item We updated the model's hyperparameters based on the validation loss and accuracy to improve the model's accuracy and ability to generalize.
	\item We used class weights during training to address class imbalance.
\end{itemize}

Table \ref{modelHyperparams} describes the network hyperparameters and table \ref{trainHyperparams} describes the training hyperparameters.

\begin{table}
	\centering
	\caption{The model's hyperparameters.}
	\label{modelHyperparams}
	\begin{tabular}{@{}lr@{}}
		\toprule
		Hyperparameter & Value \\
		\midrule
		Input Dimensions & 278w x 200h x 1d \\
		Optimization Algorithm & Adam \\
		Loss Measure & Binary Cross-entropy \\
		Accuracy Metric & Categorical Cross-entropy \\
		Activation Function & Sigmoid \\
		\bottomrule
	\end{tabular}
\end{table}

@@

The final model was evaluated on the testing set by using average accuracy. The average accuracy is calculated from the confusion matrices shown in figure \ref{confusionMatrix}. The confusion matrices are calculated by comparing the true labels to the predicted labels for each emotion. There is a confusion matrix for each emotion due to the multi-label problem considering each emotion as independent.

@@

After training the model, it was evaluated on the testing set to get the final accuracy. The final accuracy achieved was 57.64\%. Confusion matrices for each emotion are shown in Figure \ref{confusionMatrix} and comparisons to the current literature is shown in table \ref{litacccompare}.

@@

\begin{table}
	\centering
	\caption{Comparison of SER accuracy in literature.}
	\label{litacccompare}
	\begin{tabular}{@{}lrr@{}}
		\toprule
		Research Work & Method & Accuracy (\%) \\
		\midrule
		Zhao et al. \cite{Zhao2019} & CNN + LSTM & 52.1 \\
		Our Work & CNN & 57.6 \\
		Etienne et al. \cite{Etienne2018} & CNN + LSTM & 61.7 \\
		Zhang et al. \cite{Zhang2019} & CNN & 63.9 \\
		Fayek et al. \cite{Fayek2017} & CNN & 64.8 \\
		Yenigalla et al. \cite{Yenigalla2018} & CNN & 73.9 \\
		Badshah et al. \cite {Badshah2019} & CNN & 84.3 \\
		\bottomrule
	\end{tabular}
\end{table}

The results achieved in this paper do not reach the state-of-the-art accuracy achieved for the single-label case but this is the first attempt on the harder case of mutli-label using more data. So we conclude that it is possible to achieve a reasonable accuracy on the problem of multi-label speech emotion recognition by applying deep learning to the problem.

@@

One limitation of this work is that we only accounted for seven emotions but recent research has suggested that there are more emotions such as boredom, shame, and triumph. \cite{Cordaro2018} However, one issue with expanding the set of considered emotions is the lack of databases with the labeled emotion.

Another limitation of this work is that all samples are spoken in English so the model is biased towards Anglophones. In theory, the basic emotions are universal across languages and cultures so incorporating databases spoken in different languages, such as the Emo-DB database, would help the model generalize across languages.\cite{Burkhardt2005}

>>>>

The field of affective computing studies the development of systems that can recognize, interpret, process, and simulate human affects. A subfield of affective computing, the one that we are interested in, is speech emotion recognition (SER). SER is the problem of recognizing which emotions are present in speech and it is important because of its applications to
\begin{itemize}
	\item call center conversations that improve service quality,
	\item speech translation systems that are more accurate at conveying emotions,
	\item and robotic pets that are more compassionate \cite{Koolagudi2012}.
\end{itemize}

@@

SER has historically been approached in two ways:
\begin{itemize}
	\item Feature engineering
	\item Machine learning (ML)
\end{itemize}

Feature engineering is a method of manually extracting desired parameters from the data for use in  recognition. For audio, examples of parameters include pitch, energy, and Mel-Frequency Cepstral Coefficients (MFCC) \cite{Rybka2013}. Nassif et al. \cite{Nassif2019} found that most researchers used the MFCC parameter for deep learning models but also recommended other parameters such as Linear Predictive Coding.

Machine learning is a method of creating models that automatically extract their own parameters for recognition without being manually programmed. Examples include convolutional neural networks (CNN), recurrent neural networks (RNN), and long short-term memory (LSTM). Both approaches are not exclusive and a hybrid approach of feature engineering and machine learning results in better classification accuracy \cite{Nassif2019}.

@@

Current state-of-the-art SER has achieved an accuracy of 52.14\% for the case of speaker-independent, single-labeled emotions on the IEMOCAP database \cite{Zhao2019}. This result was achieved using a 2D CNN LSTM model where the audio waveform was converted into a log-Mel spectrogram and then fed into the model.

@@

The IEMOCAP \cite{busso_2008} and CREMA-D \cite{cao_2014} databases both include multiple labels for each speech sample but the data is discarded by considering the emotion with the majority of votes. We argue that discarding the non-majority votes results in a less realistic model of emotion classification due to not matching human performance and due to the loss of information. For example, in the case of speech translation systems, translating "surprise" speech lacks the nuance of whether the surprise is "surprise-happy" such as the case of receiving a gift or if it is "surprise-sad" such as the case of receiving news of the death of a loved one. A single label is not able to capture this nuance compared to multiple labels and could result in inaccurate or offensive translations.

@@

The main idea of this paper is to build a more realistic speech emotion recognition system that can recognize multiple emotions in speech. We approach this problem by using more data and by using a 2D CNN to classify speech samples into multiple emotions. We collected four databases: two with multi-label and single-label samples, and two with only single-label samples. We then combined the four databases into an a single database by processing all of the speech samples into log-Mel spectrograms. These log-Mel spectrograms were then fed into an eight-layer neural network consisting of four convolutional layers and four dense layers.

This paper is outlined as follows:
\begin{itemize}
	\item Section 2 details the methodology we use such as the how we combined the four databases and describes the architecture of the neural network.
	\item Section 3 describes and discusses the results from testing the neural network.
	\item Section 4 concludes this paper and provides suggestions for future directions.
\end{itemize}

@@

We approach the problem of SER by preprocessing samples from four databases and then feeding those samples into a CNN. We preprocess the speech samples into log-Mel spectrogram to help the neural network extract features relevant to emotion recognition. This choice is based on previous work such as \cite{Engel2019} and \cite{Badshah2017}. The choice of using a CNN is also based on previous work where Balakrishnan et al. \cite{Balakrishnan2017} found that CNN-based models have superior performance compared to RNNs. Another justification for using CNNs is due to treating the log-Mel spectrograms as images and CNNs have been shown to perform well on images \cite{Krizhevsky2012}.

A high level overview of the data flow is shown in figure \ref{highLevelDataFlowDiagram} where samples flow through three stages of processing. The stages are detailed further into the methodology. Both the preprocessing steps and the CNN were implemented in Python using the Librosa \cite{McFee2015} and Keras \cite{Chollet2015} libraries respectively.

\subsection{Preprocessing} \label{Preprocessing}

@@

To build a more realistic SER system, we first need to obtain speech samples with their labeled emotion. We considered eight databases and chose four based on accessibility and based on the number of overlapping emotions. Table \ref{dbEmoTable} compares the set of emotions of each database which makes it easier to identify overlapping emotions. We chose four databases because it becomes more difficult to maintain consistency due to database variability. Databases can vary in
\begin{itemize}
	\item The set and number of labeled emotions
	\item The number of labels per sample
	\item The audio quality such as sampling rate and noise
	\item The spoken language
\end{itemize}
Given this variability, we chose the following four databases
\begin{enumerate}
	\item IEMOCAP \cite{busso_2008}
	\item TESS \cite{dupuis_2011}
	\item RAVDESS \cite{livingstone_2018}
	\item CREMA-D \cite{cao_2014}
\end{enumerate}

@@

To control for the number of labels per sample, we mixed the multi-labeled data from the IEMOCAP and CREMA-D databases with the single-labeled data from the TESS and RAVDESS databases. We chose to mix of single- and multi- labeled data to increase the number of samples that the model can learn from and because the single-labels can be considered as special cases of multi-labeled data. However, we removed samples that were labeled with four or five emotions because we consider these samples to be ambiguous and because outliers can hinder a neural network's ability to learn. Outliers can hinder a neural network's ability to learn by causing large gradient updates that prevent the model from converging \cite{Chollet2017}.

@@

Each sample from a database went through the same preprocessing steps to maintain consistency across all samples. The detailed steps are described below.
\begin{enumerate}
	\item A sample starts as a raw waveform in the form of time series points specifying the amplitude at a point in time.
	\item The sample is then padded or cropped to the desired length of 4.5s. Shorter samples were zero-padded on the right tail. Longer samples were right-tail cropped and the extra information was discarded.
	\item If the sample came from a database that we considered noisy, then a noise reduction filter was applied to the sample. We consider the IEMOCAP and CREMA-D databases to be noisy.
	\item The sample is then converted into a log-Mel spectrogram using the STFT and Mel scale. The phase information was discarded as it does not appear to hold relevant information \cite{Kozakowski2017}.
	\item The final step is to normalize the spectrograms to have values between negative one and one. This was done by using a min-max scaling function.
\end{enumerate}

@@

The discrete-time STFT is defined in (\ref{stft}) where $x[n]$ is the discrete signal and $w[n]$ is the window function. We used a window size of 3072 with a 75\% overlap. This choice is based on the work of Zhao et al. \cite{Zhao2019} where they also use 75\% overlap but with a smaller window size (2048) and they achieved excellent results.

@@

The Mel scale is defined in (\ref{melScale}) where $f$ is the frequency in hertz and $m$ is mels \cite{Oshaughnessy1990}. We set the minimum frequency to 20 Hz and maximum frequency to 12 kHz with 200 Mel bins. The frequency range was chosen after experimenting with various ranges and selecting the range that resulted in visually clean spectrograms. The number of Mel bins was also chosen after experimentation as too few bins resulted in poor temporal resolution while too many bins resulted in poor frequency resolution.

\begin{equation}
\label{minmaxscale}
\begin{aligned}
X_{std} &= \frac{X - X_{min}}{X_{max} - X_{min}}\\
X_{scaled} &= X_{std} \cdot  (max - min) + min
\end{aligned}
\end{equation}

The min-max scaling is defined in (\ref{minmaxscale}) where $X_{min}$ is the smallest value in tensor $X$, $X_{max}$ is the largest value in tensor $X$, $min$ is the lower bound set to $-1$, and $max$ is the upper bound set to $1$. We scaled all spectrograms to only have values between $-1$ and $1$ as this reduces the chances of triggering large gradient updates that prevent the model from converging \cite{Chollet2017}.

@@

\begin{figure*}
	\centering
	\hspace{6mm}
	\includegraphics[width=\textwidth]{neural_network_architecture.png}
	\caption{Architecture for the SER network including the input layer and eight processing layers.}
	\label{neuralNetworkArchitecture}
\end{figure*}

After all of the databases were processed as described in section \ref{Preprocessing}, the final combined database was fed into a neural network for training, validation, and testing. We constructed an eight layer neural network consisting of four convolutional layers and four dense layers. The network architecture is shown in figure \ref{neuralNetworkArchitecture}. The training process is described below.
\begin{itemize}
	\item We shuffled the combined database to make each input batch more uniform thus mitigating large gradient updates.
	\item We split the combined database into 80\% training, 10\% validation, and 10\% testing.
	\item We applied dropout to the dense layers and batch normalization to the convolutional layers to deal with overfitting \cite{Srivastava2014}, \cite{Ioffe2015}.
	\item We updated the model's hyperparameters based on the validation loss and accuracy to improve the model's accuracy and ability to generalize.
	\item We used class weights during training to address class imbalance.
\end{itemize}

Table \ref{modelHyperparams} describes the model's hyperparameters and table \ref{trainHyperparams} describes the training hyperparameters. The decision to use \textit{binary cross-entropy} as the loss measure is due to multi-label problems considering each output as independent and thus each output acts like a binary classifier. This lets the model only penalize the outputs that are incorrect without penalizing the other outputs. However, we use \textit{categorical cross-entropy} as the accuracy metric because binary cross-entropy would provide the incorrect accuracy as it defined for a binary output such as 0 or 1, not vectors containing binary outputs. We could have picked a different accuracy metric but chose categorical cross-entropy due to its ease of use.

Likewise, we use \textit{sigmoid} as the output layer's activation function instead of the more commonly used \textit{softmax} function in classification. The justification is the same as in our choice of loss measure; sigmoid treats each output as independent which means multiple outputs can be true while softmax maximizes a single output while minimize all other outputs.

\begin{table}
	\centering
	\caption{The model's hyperparameters.}
	\label{modelHyperparams}
	\begin{tabular}{@{}lr@{}}
		\toprule
		Hyperparameter & Value \\
		\midrule
		Input Dimensions & 278w x 200h x 1d \\
		Optimization Algorithm & Adam \\
		Learning Rate & 0.001 \\
		Beta 1 & 0.9 \\
		Beta 2 & 0.999 \\
		Epsilon & $10^{-8}$ \\
		Loss Measure & Binary Cross-entropy \\
		Accuracy Metric & Categorical Cross-entropy \\
		Hidden Layer Activation Function & ReLU \\
		Output Layer Activation Function & Sigmoid \\
		\bottomrule
	\end{tabular}
\end{table}

@@

The final model was evaluated on the testing set by using average accuracy that was calculated from the confusion matrices shown in figure \ref{confusionMatrix}. The confusion matrices are calculated by comparing the true labels to the predicted labels for each independent emotion.

@@

After training the model, it was evaluated on the testing set to get the final accuracy. The final accuracy achieved was 57.64\%. Confusion matrices for each emotion are shown in Figure \ref{confusionMatrix} and comparisons to the current literature is shown in table \ref{litacccompare}. We compare our results to the single-label case of SER as we could not find papers with results for the multi-label case.

@@

\begin{table*}
	\centering
	\caption{Comparison of SER accuracy in literature.}
	\label{litacccompare}
	\begin{tabular}{@{}llllllr@{}}
		\toprule
		Research Work & Method & Testing Method & Number of Emotions & Databases Used & Label Type & Accuracy (\%) \\
		\midrule
		Zhao et al. \cite{Zhao2019} & CNN + LSTM & Testing set & 6 & IEMOCAP & Single & 52.1 \\
		Our Work & CNN & Testing set & 7 & \begin{tabular}[c]{@{}l@{}}IEMOCAP, TESS, \\ RAVDESS, CREMA-D\end{tabular} & Multi & 57.6 \\
		Etienne et al. \cite{Etienne2018} & CNN + LSTM & Testing set & 4 & IEMOCAP & Single & 61.7 \\
		Zhang et al. \cite{Zhang2019} & CNN & Testing set & 4 & IEMOCAP & Single & 63.9 \\
		Fayek et al. \cite{Fayek2017} & CNN & Testing set & 4 & IEMOCAP & Single & 64.8 \\
		Yenigalla et al. \cite{Yenigalla2018} & CNN & Testing set & 4 & IEMOCAP & Single & 73.9 \\
		Badshah et al. \cite {Badshah2019} & CNN & Testing set & 7 & Emo-DB, Korean dataset & Single & 80.8 \\
		\bottomrule
	\end{tabular}
\end{table*}

The results achieved in this paper do not reach the state-of-the-art accuracy achieved for the single-label case. But considering how this is the first attempt on the more realistic problem of multi-label SER, we perform reasonably well. Compared to the literature shown in table \ref{litacccompare}, no other paper considers as many emotions as we do while also using four databases and also tackling the multi-label problem. So we conclude that we are successful in building a more realistic SER system by using ML and more data to tackle the problem of recognizing multiple emotions in speech.

@@

One limitation of this work is that we only accounted for seven emotions but recent research has suggested that there are more emotions such as boredom, shame, and triumph \cite{Cordaro2018}. However, one issue with expanding the set of considered emotions is the lack of databases with the labeled emotion.

Another limitation of this work is that all samples are spoken in English so the model is biased towards Anglophones. In theory, the basic emotions are universal across languages and cultures so incorporating databases spoken in different languages, such as the Emo-DB database, would help the model generalize across languages \cite{Burkhardt2005}.
